---
title: "session 2 flat files"
author: "Tim Riffe"
date: "2022-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# flat files

Flat files are read in using metadata on columns positions and widths. That's why we download the pdf manual. If they were polite they would give a spreadsheet of column positions and widths, but they deliver it in a pdf to be jerks.

```{r}
library(readr)

fwf_metadata <- fwf_positions(start = c(19,21,75),
                              end = c(20,22,76), 
                              col_names = c("hour","minute","mage"))

NAT <- read_fwf("Data/nat2021us.txt.gz", 
                col_positions = fwf_metadata)
# examine this knarly data to get a sense of it.
#readLines("Data/nat2021us.txt.gz",n = 10)


NAT %>% 
  group_by(mage,hour,minute) %>% 
  summarize(births = n(),
            .groups = "drop") %>% 
  mutate(time = as.POSIXct(
                  paste(hour, 
                        minute, 
                        sep = ":"),
                  format = "%H:%M")) %>% 
  ggplot(aes(x = time, y = mage, fill = births)) +
  geom_tile()
```

Exercise:

1. Instead of births, plot the fraction of births at each time of day for each mother age.

2. plot it as a ridge plot. Hahahaha. Maybe every 5th mother age? And maybe just 15-45?

3. break it down by cesarean status. Meaning that the distributions should be by both age and cesariean status. Plot these side-by-side in a ridge plot just like we did for asfr before.






