---
title: "Session 1 Notes"
author: "Tim Riffe"
date: "2022-11-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# a bit about R markdown

To make a code chunk, press `Ctrl Alt i` 
```{r}
# this bit in here is R
```


Because it was asked (HT Ariane)
To comment something out, select it and press
`Ctrl Shift c`

Observe this

<!-- You won't see this -->

# Let's start talking about tidy data

Tidy data is a data where observations are in rows and variables are in columns. It's a rectangular data format.

Why? because it's easy to model (stats) and easy to map (data viz).

# `gapminder` data is tidy

Let's give an example of tidy data. `str()` gives lots of good metadata.
```{r}
# install.packages("gapminder")
library(gapminder)
head(gapminder)
str(gapminder)
```

Observations here are unique combinations of country and year.

You can make a pip with Ctrl Shift m.

```{r}
library(janitor)
library(tidyverse, quietly = TRUE)
# install.packages("tidyverse")
clean_names(gapminder)

# this is identical, to be explained later
gapminder %>% 
  clean_names()

gapminder %>% 
  rename(le = lifeExp,
         gdp = gdpPercap)
```

`clean_names()` turns names into so-called snake-case, which is easier to type fast than camel case. But it's a strict output. You can do custom names with `rename()`, which follows the pattern `new = old` when renaming variables.

# filtering and selecting


```{r}
# g1 <-
gapminder %>% 
  clean_names() %>% 
  filter(life_exp > 80) #-> g1

# identical to 
# range(gapminder$gdpPercap)

gapminder %>% 
  pull(gdpPercap) %>% 
  range()

gapminder %>% 
  clean_names() %>% 
  filter(between(life_exp, 50, 80),
         gdp_percap > 5000)
```

Filtering can be one one or more logical condition, referring to columns in the data or not. If there is more than one condition, then these can either be `&`-separated, i.e. `x & y & z`, or comma-separated (which I prefer because it allows nicer line-breaks) like so `filter(x,y,z)`. There are helper functions for these equality tests, to see a list of them, look at `?between`. `&`, `|`, `>`,`>=`, `==`, `<`,`<=`. `any()`, `all()`, `!`, `!=`, `%in%`.

Set membership is a neat one, `%in%`.
```{r}
gapminder %>% 
  filter(country %in% c("Afghanistan","Burundi","Jamaica"))


gapminder_unfiltered %>% 
  clean_names() %>% 
  filter(grepl(country, pattern = "stan")) %>% 
  pull(country) %>% 
  as.character() %>% 
  unique()

```

Selecting is for picking out columns. You can do so positively, negatively, by name, by position, for multiple things, etc.
```{r}
gapminder %>% 
  clean_names() %>% 
  select(country, life_exp)

gapminder %>% 
  clean_names() %>% 
  select(-continent)

gapminder %>% 
  select(starts_with("c"))

gapminder %>% 
  clean_names() %>% 
  select(-pop, -gdp_percap)
```

Concepts you've seen so far:
`filter()` , `select()`, `pull()`, 
and helpers:
`starts_with()`, `between()`, `contains()` bla bla

`mutate()`, `group_by()`, `summarize()`,`pivot_longer()`, `pivot_wider()`.

# `mutate()`

`mutate()` does things to columns. It changes a column, or it makes a new one. It does *not* change the rows. You can do multiple operations in a `mutate()` call, comma separated, and they can have sequential dependencies, meaning you can use things that you just made.
```{r}
gapminder %>% 
  clean_names() %>% 
  mutate(gdp = pop * gdp_percap,
         gdp_lifetime = gdp_percap * life_exp,
         gdp_100 = 100 * gdp_percap,
         lgdp = log(gdp),
         big_or_rich = gdp > 1e11)
```

You can also calculate variables for groups, while staying in the original row format of the data, i.e. without shrinking it down to just the group stats. Observe: Let's calculate the decade avg gdp_percap.

```{r}
0:10 - 0:10 %% 5
0:10 %% 5
gapminder %>% 
  clean_names() %>% 
  mutate(decade = year - year %% 10) %>% 
  group_by(country, decade) %>% 
  mutate(gdp_percap_10 = mean(gdp_percap),
         gdp_percap_10w = sum(gdp_percap * pop) / sum(pop))
```


Groups are strict- not just for calculating, but also for filtering and anything else we're talking about now, see:
```{r}
gapminder %>% 
  clean_names() %>% 
  group_by(year) %>%  # this is strict!
  filter(life_exp == max(life_exp)) %>% 
  ungroup() %>% 
  arrange(year)

gapminder %>% 
  arrange(country,-year)
```

# `summarize()`

`summarize()` is different from `mutate()` in that it can change the rows, and we usually use it for aggregation. And that usually happens within groups.
```{r}
gapminder %>% 
  clean_names() %>% 
  group_by(country) %>% 
  summarize(gdp_cumulative = sum(gdp_percap * pop) * 5) %>% 
  ungroup() %>% 
  arrange(-gdp_cumulative)
```

Above we calculate the total gdp over the second half of the 20th century. Note we scale the 5-year groups in the sum, so that it's a proper annual amount being summed. Assuming a step function because we're lazy.

Here we demonstrate the same sort of aggregation, aiming to calculate the global gdp using the unfiltered version of the gapminder data, but upon visualizing it we see some irregularities that likely are due to some countries having annual interpolations and others not.
```{r}
gapminder_unfiltered %>% 
  clean_names() %>% 
  group_by(year) %>% 
  summarize(gdp_global = sum(gdp_percap * pop)) %>% 
  
  ggplot(aes(x = year, y = gdp_global)) +
  geom_line() + 
  geom_smooth(method = 'lm')
```

Which countries only have 5-year data?
```{r}
gapminder_unfiltered %>%
  clean_names() %>% 
  group_by(country) %>% 
  filter(!any(year == 1951)) %>% 
  pull(country) %>% 
  as.character() %>% 
  unique()
```

We conclude that Hans Rosling took HMD data for some countries and some WPP selection for the others.

# pivoting

To make data wider, use `pivot_wider()`. Names get concatenated
```{r}
gapminder %>% 
  clean_names() %>% 
  pivot_wider(names_from = year,
              values_from = c(pop, life_exp, gdp_percap))
```

To make data longer use `pivot_longer()`

```{r}
gapminder %>% 
  clean_names() %>% 
  pivot_longer(c(pop, life_exp, gdp_percap),
               names_to = "variable",
               values_to = "value") %>% 
  pivot_wider(names_from = variable,
              values_from = value)
```

A more complicated example getting from wide back to long / tidy. In this case it's a little tricky because names have underscores in them sometimes, so we specify a different separator (`":"`), then we pivot to *very long*, separate columns on the `:` and then pivot back to the original form. That's convoluted, I admit. But, let's try to swallow that sometimes data wrangling makes us do uncomfortable gymnastics for the sake of getting the job done. And you should know that there is no job that you cannot do. There is no dataset that can defeat you.

```{r}
gapminder %>% 
  clean_names() %>% 
  pivot_wider(names_from = year, 
              values_from = c(pop, life_exp, gdp_percap),
              names_sep = ":") %>% 
  pivot_longer(-c(1,2),
               names_to = "var_year",
               values_to = "value") %>% 
  separate(col = var_year, into = c("variable","year"), sep = ":") %>% 
  pivot_wider(names_from = variable,
              values_from = value)
```



















