---
title: "session 2 notes"
author: "Tim Riffe"
date: "2022-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r, eval = FALSE}
url_download_trigger_thingy <-
"http://data.un.org/Handlers/DocumentDownloadHandler.ashx?id=319&t=bin"

options(timeout = 200)
download.file(
  url = url_download_trigger_thingy,
  destfile = "Data/un_fertility.xls"
)

```

Read in the data using `readxl`

```{r}
library(tidyverse)
library(readxl)
library(janitor)
ASFR <- read_excel(path = "Data/un_fertility.xls",
           skip = 4,
           na = "..") %>% 
  select(country = "...1",
         LocID = "...2",
         Year,
         "15" = "15-19",
         "20" = "20-24",
         "25" = "25-29",
         "30" = "30-34",
         "35" = "35-39",
         "40" = "40-44",
         "45" = "45-49"
         ) %>% 
  pivot_longer(`15`:`45`,
               # 4:10
               names_to = "age",
               values_to = "asfr") %>% 
  clean_names() %>% 
  mutate(asfr = asfr / 1000)
```

Now we have read in the fertility rates, and standardized names and reshaped to a workable format, so they are tidy but not clean, boo. That's because there are anomalous cases in the data where synthetic indicators are available but no age-specific rates. So we begin by filtering out those cases.

```{r}
ASFR <-
ASFR %>% 
  filter(! is.na(asfr)) %>% 
  group_by(country, year) %>% 
  #mutate(n = n()) %>% 
  # keep only subsets where we 
  # have at least 5 age groups
  filter(n() >= 5) %>% 
  mutate(age = as.integer(age)) %>% 
  ungroup()

```


We make a custom graduation function based on running a monotonic spline through the cumulative fertility rate curve, taking care to scale up the rates in the cumulative distribution and align them properly with the hard right side of each age group. Further, we anchor the endpoints with fertility rates of 0 at some declared ages. Then we predict for single ages within the original range of the data.
```{r}

my_graduatation_hack <- function(chunk){
  x <- chunk$age
  y <- chunk$asfr 
  
  Y <- cumsum(c(0,0,y,0,0)) * 5
  X <- c(13,14,x+5,55,56)
  
  y_out <-
     splinefun(x = X,
               y = Y, 
               method = "monoH.FC")(14:49) %>% 
     diff()
  
  chunk_out <- tibble(age = 15:49,
                      asfr = y_out)
  return(chunk_out)
}
# example usage
chunk <- ASFR %>% 
  filter(loc_id == 4,
         year == 1973)
chunk_out <- my_graduatation_hack(chunk)

plot(chunk$age, chunk$asfr, type = 's')
lines(chunk_out$age, chunk_out$asfr)
```

Now we can apply this function for all subsets after a `group_by()`, and note that what we include in the grouping variables we also keep.

```{r}
ASFR1 <-
  ASFR %>% 
  group_by(country, year, loc_id) %>% 
  group_modify(~ my_graduatation_hack(chunk = .x)) %>% 
  ungroup()
```

Now we aim to plot for each TFR level the distribution that belongs to the highest and lowest mean age at childbearing.

Weighted mean general formula
$$
\bar{x} = \frac{\sum x \cdot w}{\sum w}
$$

```{r}
asfr_extremes <-
ASFR1 %>% 
  group_by(country, year, loc_id) %>% 
  mutate(tfr = sum(asfr),
         mab = sum((age+.5) * asfr) / sum(asfr)) %>% 
 ungroup() %>% 
 mutate(tfr_interval = tfr - tfr %% .2) %>% 
 group_by(tfr_interval) %>% 
 #filter(mab %in% range(mab))
 mutate(extreme = case_when(
                    mab == max(mab) ~ "max",
                    mab == min(mab) ~ "min",
                    TRUE ~ NA_character_
 )) %>% 
 ungroup() %>% 
 filter(!is.na(extreme))
```

mini explanation of `case_when()`. You list conditions (meaning that you need expressions that evaluate to logical, meaning they return TRUE or FALSE), in order from the most specific to the most general. All cases must be handled. You can close it out with TRUE, because TRUE evaluates to TRUE meaning "everything else".

Now let's make the ridge plot to compare distributions.
```{r}
library(ggridges)

asfr_extremes %>% 
  filter(between(tfr, 1.2, 6)) %>% 
  ggplot(mapping = aes(x = age,
                       y = factor(tfr_interval),
                       height = asfr,
                       fill = extreme)) +
  geom_ridgeline(alpha = .5,
                 scale = 2)

```



